### 2016-01-14 model/overflowmcmc.tex
Fixed a couple typos

In order for the gamma process to be well defined we require $\phi>0$ and $\Psi(\tau)$ to be nondecreasing and right continuous with $\Psi(0)=0$. An easy way to achieve this is to define $\Psi(\tau) = \int_0^\tau\psi(t)dt$ where $\psi(t)$ is continuous and nonnegative. While $\psi(t)$ can be thought of as the intensity of the Gamma process it is also intimately related to the intensity of the Poission process. The mean of $N(\tau + \delta) - N(\tau)$ is 
\begin{align*}
E[N(\tau + \delta) - N(\tau)] &= E\{E[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\} = E[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
 &= [\Psi(\tau + \delta) - \Psi(\delta)]/\phi = \int_{\tau}^{\tau + \delta}\psi(t)dt/\phi.
\end{align*}
In this way, $\psi(t)/\phi$ can be seen as the expected intensity function of the $N(\tau)$ process. The Gamma-Poisson structure also allows for a more flexible model on $N(\tau)$ by allowing for overdispersion. In a Poisson model the mean and variance are the same, but in this case they are proportional:
\begin{align*}
V[N(\tau + \delta) - N(\tau)] &= E\{V[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\} + V\{E[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\}\\
&= E[\Lambda(\tau + \delta) - \Lambda(\tau)] + V[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
&= [\Psi(\tau + \delta) - \Psi(\tau)](1/\phi + 1/\phi^2) = \frac{\phi + 1}{\phi}E[N(\tau + \delta) - N(\tau)].
\end{align*}
This allows for overdispersion but not underdispersion since $(\phi + 1)/\phi\in[1,\infty)$ for $\phi\ge 0$.

@@

\subsection{Block arrival times --- $\tau_k$}
The full conditional distribution of the block arrival times is more complicated:
\begin{align*}
p(\bm{\tau}_{1:n_{b}}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\frac{\phi^{\Psi(\tau_k) - \Psi(\tau_{k-1})}}{\Gamma\left[\Psi(\tau_k) - \Psi(\tau_{k-1})\right]}\lambda_k^{\Psi(\tau_k)-\Psi(\tau_{k-1}) - 1}
\end{align*}
where $\tau_0\equiv0$ and $\Gamma(x) = \int_0^\infty t^{x-1}e^{-t}dt$ is the gamma function. Even if $\Psi(\tau)$ is some nice known form, this density is nonstandard because the $\tau_k$'s enter the Gamma function. Some ideas for handlilng this step:
\begin{enumerate}
\item Draw each $\tau_k$ condtional on all the others using random walk Metropolis steps for $k=1,\dots,n_{b}$.
\item Approximate the density of $\tau_k$ sufficiently well and use independent Metropolis steps for $k=1,\dots,n_{b}$.
\item Approximate the density of $\tau_{1:n_{b}}$ sufficiently well and use an independent Metropolis step.
\item Use adaptive rejection sampling (ARS) to draw from the density of $\tau_k$.
\item Use ARMS --- ARS except with a Metropolis correction for densities which are not log concave (look it up).
\end{enumerate}

\subsection{Transaction process dispersion --- $\phi$}
The full conditional density of $\phi$ is
\begin{align*}
p(\phi|...) \propto \prod_{k=1}^{n_{b}}\phi^{\psi_k}e^{\lambda_k\phi}p(\phi) \propto \phi^{\sum_{k=1}^{n_{b}}\psi_k}e^{\phi\sum_{k=1}^{n_{b}}\lambda_k}p(\phi).
\end{align*}
With $\phi \sim G(a_\phi, b_\phi)$ in the prior, this is a Gamma density so the step is simple:
\begin{enumerate}
\item Draw $\phi \sim G\left(a_\phi + \sum_{k=1}^{n_{b}}\psi_k, b_{\phi} + \sum_{k=1}^{n_{b}}\lambda_k \right)$.
\end{enumerate}

>>>>

In order for the gamma process to be well defined we require $\phi>0$ and $\Psi(\tau)$ to be nondecreasing and right continuous with $\Psi(0)=0$. An easy way to achieve this is to define $\Psi(\tau) = \int_0^\tau\psi(t)dt$ where $\psi(t)$ is continuous and nonnegative. While $\psi(t)$ can be thought of as the intensity of the gamma process it is also intimately related to the intensity of the Poission process. The mean of $N(\tau + \delta) - N(\tau)$ is 
\begin{align*}
E[N(\tau + \delta) - N(\tau)] &= E\{E[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\} = E[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
 &= [\Psi(\tau + \delta) - \Psi(\delta)]/\phi = \int_{\tau}^{\tau + \delta}\psi(t)dt/\phi.
\end{align*}
In this way, $\psi(t)/\phi$ can be seen as the expected intensity function of the $N(\tau)$ process. The gamma-Poisson structure also allows for a more flexible model on $N(\tau)$ by allowing for overdispersion. In a Poisson model the mean and variance are the same, but in this case they are proportional:
\begin{align*}
V[N(\tau + \delta) - N(\tau)] &= E\{V[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\} + V\{E[N(\tau + \delta) - N(\tau)|\Lambda(\tau + \delta) - \Lambda(\tau)]\}\\
&= E[\Lambda(\tau + \delta) - \Lambda(\tau)] + V[\Lambda(\tau + \delta) - \Lambda(\tau)]\\
&= [\Psi(\tau + \delta) - \Psi(\tau)](1/\phi + 1/\phi^2) = \frac{\phi + 1}{\phi}E[N(\tau + \delta) - N(\tau)].
\end{align*}
This allows for overdispersion but not underdispersion since $(\phi + 1)/\phi\in[1,\infty)$ for $\phi\ge 0$.

@@

\subsection{Block arrival times --- $\tau_k$}
The full conditional distribution of the block arrival times is more complicated:
\begin{align*}
p(\bm{\tau}_{1:n_{b}}|...) \propto \exp\left[-\frac{1}{2\sigma^2}\sum_{k=1}^{n_{b}}\sum_{i=1}^{\tilde{n}_k+1}\left(\tilde{t}_{ik} - \tau_k\right)^2 - \frac{1}{10}\tau_{n_{b}}\right]\prod_{k=1}^{n_b}\frac{\phi^{\Psi(\tau_k) - \Psi(\tau_{k-1})}}{\Gamma\left[\Psi(\tau_k) - \Psi(\tau_{k-1})\right]}\lambda_k^{\Psi(\tau_k)-\Psi(\tau_{k-1}) - 1}
\end{align*}
where $\tau_0\equiv0$ and $\Gamma(x) = \int_0^\infty t^{x-1}e^{-t}dt$ is the gamma function. Even if $\Psi(\tau)$ is some nice known form, this density is nonstandard because the $\tau_k$'s enter the gamma function. Some ideas for handling this step:
\begin{enumerate}
\item Draw each $\tau_k$ condtional on all the others using random walk Metropolis steps for $k=1,\dots,n_{b}$.
\item Approximate the density of $\tau_k$ sufficiently well and use independent Metropolis steps for $k=1,\dots,n_{b}$.
\item Approximate the density of $\tau_{1:n_{b}}$ sufficiently well and use an independent Metropolis step.
\item Use adaptive rejection sampling (ARS) to draw from the density of $\tau_k$.
\item Use ARMS --- ARS except with a Metropolis correction for densities which are not log concave (look it up).
\end{enumerate}

\subsection{Transaction process dispersion --- $\phi$}
The full conditional density of $\phi$ is
\begin{align*}
p(\phi|...) \propto \prod_{k=1}^{n_{b}}\phi^{\psi_k}e^{\lambda_k\phi}p(\phi) \propto \phi^{\sum_{k=1}^{n_{b}}\psi_k}e^{\phi\sum_{k=1}^{n_{b}}\lambda_k}p(\phi).
\end{align*}
With $\phi \sim G(a_\phi, b_\phi)$ in the prior, this is a gamma density so the step is simple:
\begin{enumerate}
\item Draw $\phi \sim G\left(a_\phi + \sum_{k=1}^{n_{b}}\psi_k, b_{\phi} + \sum_{k=1}^{n_{b}}\lambda_k \right)$.
\end{enumerate}
